Recovery of mountain plant communities in response to reductions in Nitrogen emissions is hidden by other drivers of global change

### 2018-06-15 Manuscript.Rmd
Ologotrophic species richness ergänzt. Diskussion überarbeitet.

# Estimate effect of N deposition at a given time point
makemod <- function(year) {
  mod <- glm(SR ~ Temperature + Precipitation + NTOT + Inclination,
             data = dat %>% filter(yearP >= year & yearP <= (year + 4)), family = poisson)
  arm::sim(mod, nsim)@coef
}

@@

# Discussion {-}
Although Nitrogen deposition considerably declined between 2000 and 2015 (Fig. \@ref(fig:Ndepchange)), we could not detect major shifts in plant community structure during the same time period (Table \@ref(tab:communitytrendstab)). While the slight increase in average temperature value suggests that plant communities adopted to increasing temperature, the constant aveage nutrients value suggests that the decrease in Nitrogen deposition did not yet affected plant communities. However, the stability in community structure was accompanied by marked turnover in species identities. Species turnover especially varied along the Nitrogen deposition gradient with highest species turnover at sites with low Nitrogen deposition. These results are in line with recent studies suggesting that species loss (or gain) is not necessarily apparent in assessments of contemporary communities [@Vellend2013, @Dornelas2014, @Hillebrand2018].

Species communities are shaped by a range of factors, including deterministic processes such as environmental filtering or competitive interactions [@Goetzenberger2012, @Janevcek2013]. Such factors select for species with specific characteristics. Community assembly theories thus suggest that the factors that drive change in species communitis could be inferred from comparing the charactereistics of disappearing or colonizign species with random expectation [@Chase2011]. We adopted this idea and compared the indicator values of species that disappeared or colonized a site with the indicator values from randomly chosen species from the same site. The nutrient values of colonization species showed larges deviations from random expectation suggesting that Nitrogen deposition or other factors that change the nutrient content of the soils were the most important drivers of the change in species composition over the last 15 years in Swiss mountain hay meadows.

Nitrogen deposition at our plots decreased by `r mean(sites$NTOT_2015 - sites$NTOT_2000)` kg ha$^{-1}$ yr$^{-1}$ on average between 2000 and 2015. This is only about one tenth of the decrease in England in the same period, where Nitrogen deposition decreased by 24 kg ha$^{-1}$ yr$^{-1}$ from 1996 to 2011 [@Storkey2015]. This might explain why we observed no change in average nutrient value of communities. An other explanation might be that communtiy change due to the change in the rate of Nitrogen deposition was overruled by other anthropogenic pressures including climate change. In this study eutrophic species had rather high local survival across the entire deposition gradient, while oligotrophic species have much reduced local survival at high N deposition. This suggests that it takes more time to replace eutrophic by oligotrophic species than replacing oligotrophic by eutrophic species. This is in contrast to the species' indicator values for temperature: species with low and high temperature values had both optimal temperatures were colonization and local survival probabilities were high (see Fig. . Thus comunitiy change in response to climate warming might be fast because low temperature species have lower local survival and high temperature species have higher colonization probability.

Observational studies along a gradient of Nitrogen deposition often infer how the spatial variation in species richness is related to Nitrogen deposition. They assume that the spatial variation in species richness arose because over time some areas lost more species than others because they chronically experienced higher Nitrogen deposition. Although there is evidence supporting the use of such a 'space for time substitution' for detecting the effects of Nitrogen deposition on plant diversity [@Stevens2010], they can not replace studies that relate temporal patterns in species with Nitrogen deposition [@deSchrijver2011]. There are only a limited number of studies inferring temporal trends of plant species diversity related to varying amounts of N-deposition [@Storkey2015]. In @Clark2008, the recorded recovery in species richness within one or two decade was likely due to the species rich vegetation surrounding the experimental plots, from where immigration was easily feasible. 
This is in line with our results sugggesting that it is mainly colonizing and not disappearing species are important to shift back from euttrophic to oligotrophic communities.

- *Methological issue*: The rather large spatial turnover could be partly explained by species that remained undetected in one of the surveys. However, our results suggest that turnover is caused at least partly by species with specific indicator values. This deviation from what we would expect under random species turnover is unlikely to be explained by species that remained undetected.

- *Empirical critical loads*: Our data on colonization and local survival (i.e. temporal variation) confirm the empirical critical loads that we inferred from analysing spatial co-variation of N deposition and species richness.

>>>>

# Estimate effect of N deposition at a given time point
makemod <- function(year) {
  mod <- glm(SR_olig ~ Temperature + Precipitation + NTOT + Inclination,
             data = dat %>% filter(yearP >= year & yearP <= (year + 4)), family = poisson)
  arm::sim(mod, nsim)@coef
}

@@

# Discussion {-}
Although Nitrogen deposition considerably declined between 2000 and 2015 (Fig. \@ref(fig:Ndepchange)), we could detect only weak shifts in plant community structure during the same time period (Table \@ref(tab:communitytrendstab). While the slight increase in average temperature value suggests that plant communities adopted to increasing temperatures, the constant aveage nutrients value suggests that the decrease in Nitrogen deposition did not yet affected plant communities. However, this apparent stability in community structure was accompanied by marked temporal turnover in species identities. It seems unlikely that this temporal turnover can entirly be explained by methodological issues such as overlooked species. First, pseudo-turnover of species entities in two independent surveys of the same site during the same season was smaller than the observed temporal turnover between two surveys from different years. Second, spatial variation of turnover showed patterns that can hardly be explained by methodological issues. For instance species turnover varied along the Nitrogen deposition gradient with highest species turnover at sites with low Nitrogen deposition. Taken together, our results adds to the increasing evidence that contemporary communities are often relatively stability at community level, but this apparent stability is often accompanied by marked turnover of species [@Vellend2013, @Dornelas2014, @Hillebrand2018].

Species communities are shaped by a range of factors, including deterministic processes such as environmental filtering or competitive interactions [@Goetzenberger2012, @Janevcek2013]. Such factors select for species with specific characteristics. Community assembly theories thus suggest that the factors that drive change in species communities could be inferred from comparing the charactereistics of disappearing or colonizing species with random expectation [@Chase2011]. We adopted this idea and compared the indicator values of species that disappeared or colonized a site with the indicator values from randomly chosen species from the same site. We found that the nutrient values of colonization species showed largest deviations from random expectation suggesting that Nitrogen deposition or other factors that change the nutrient content of soils were the most important drivers of the change in species composition over the last 15 years in Swiss mountain hay meadows.

Nitrogen deposition at our plots decreased by `r mean(sites$NTOT_2015 - sites$NTOT_2000)` kg ha$^{-1}$ yr$^{-1}$ on average between 2000 and 2015. This is only about one tenth of the decrease in England, where Nitrogen deposition decreased by 24 kg ha$^{-1}$ yr$^{-1}$ from 1996 to 2011 [@Storkey2015]. This comparatively low change in Nitrogen deposition might explain why we observed no change in average nutrient value of communities. Additionally, other anthropogenic pressures such as climate change might overrule Nitrogen depositin effects on community structure. Particularly, we found that the species disappearing from the sites tend to have below average indicator values for temperature, while species that newly colonize sites have above average indicator values for temperature (Table \@ref(tab:difffromrandomtab)). Thus, the effect of disappearing and the effect of colonizating species on the community mean for temperature is additive resulting in increasing average temperature values (Tab. \@ref(tab:communitytrendstab)). This is in contrast to how Nitrogen deposition is affecting disappearing and colonization of species: It seems that both species disappearing from the sites as well as species colonizing the sites tend to have below average indicator values for nutrients (Table \@ref(tab:difffromrandomtab)). Thus, the effect of disappearing and colonizating species on the average community value for nutrients partially cancel each other out. Furthermore, eutrophic species had rather high local survival across the entire deposition gradient, while oligotrophic species have much reduced local survival at high Nitrogen deposition. This suggests that it takes more time to replace eutrophic by oligotrophic species than replacing oligotrophic by eutrophic species. Taken toghether, these reasons might explain why the response of mountain hay medow communities responded stronger to to climate warming affects than to Nitrogen reduction, altough the reduction in Nitrogen resulted in above average colonization of oligotrophic species.

Observational studies along a gradient of Nitrogen deposition often infer how the spatial variation in species richness is related to Nitrogen deposition. They assume that the spatial variation in species richness (or other metrics of community structure) arose because over time some areas lost more species than others because they chronically experienced higher Nitrogen deposition. Although there is evidence supporting the use of such a 'space for time substitution' for detecting the effects of Nitrogen deposition on plant diversity [@Stevens2010], they can not replace studies that relate temporal patterns in species with Nitrogen deposition [@deSchrijver2011]. There are only a limited number of studies inferring temporal trends of plant species diversity related to varying amounts of N-deposition [@Clark2008, @Storkey2015]. In an recent study, we used the mountain hay meadow data from a single survey only and estimated the critical load along the Nitrogen deposition gradient at which species richnes of oligotrophic species richness start to decrease with increasing Nitrogen deposition [@Roth2017]. Using this spatial variation in species richness and nitrogen deposition we estimated a critical load for mountain hay meadows of 13.1 kg ha$^{-1}$ yr$^{-1}$. In the current study we estimated the rate of Nitrogen deposition at which local survival probability or colonization probability was equal for oligotrophic and eutrophic species. This rate could also be interpreted as a critical load and revealed very similar results as the spatial approach. Our results are thus validating the space for time approach.

Hower, the results of Fig. \@ref(fig:) also shows that the results of spatial comparisons must be interpreted carefully. When we used the same model to infer how species richness is decreasing along the Nitrogen deposition gradient at differnt time points, except for the first two study periods the decrease became steeper (more negative) over time. Our first interpretation was that the Nitrogen deposition effect became stronger over time. This contraticted our prediction that the nitrogen deposition should become weaker over time since Nitrogen deposition decreased. However, we then realized that species turnover was highest in low deposition sites (Tab. \@ref(tab:turnovertab)) where colonizationg species were species with below average indicator values for Nutrients. It seems that the decrease in Nitrogen deposition resulted in higher numbers of oligotrophic species at sites with low deposition, thus leading to a steeper decline in oligotrophic species richness from sites with low to sites with high Nitrogen deposition.
